Wallace Session 2024-12-05
================

Please find below the R code history from your *Wallace* v2.1.3 session.

You can reproduce your session results by running this R Markdown file
in RStudio.

Each code block is called a “chunk”, and you can run them either
one-by-one or all at once by choosing an option in the “Run” menu at the
top-right corner of the “Source” pane in RStudio.

For more detailed information see <http://rmarkdown.rstudio.com>).

### Package installation

Wallace uses the following R packages that must be installed and loaded
before starting.

```{r}
library(spocc)
library(spThin)
library(dismo)
library(sf)
library(ENMeval)
library(wallace)
```

The *Wallace* session code .Rmd file is composed of a chain of module
functions that are internal to *Wallace*. Each of these functions
corresponds to a single module that the user ran during the session. To
see the internal code for these module functions, click on the links in
the .Rmd file. Users are encouraged to write custom code in the .Rmd
directly to modify their analysis, and even modify the module function
code to further customize. To see the source code for any module
function, just type its name into the R console and press Return.

```{r}
# example:
# just type the function name and press Return to see its source code
# paste this code into a new script to edit it
occs_queryDb
```

Your analyses are below.

------------------------------------------------------------------------

## Analysis for *Mephitis mephitis* (Mm)

User CSV path with occurrence data. If the CSV file is not in the
current workspace, change to the correct file path
(e.g. “/Users/darwin/Documents/occs/”).

```{r}
# NOTE: provide the folder path of the .csv file
occs_path <- ""
occs_path <- file.path(occs_path, "striped_skunk2.csv")
# get a list of species occurrence data
userOccs_Mm <- occs_userOccs(
  txtPath = occs_path, 
  txtName = "striped_skunk2.csv", 
  txtSep = ",", 
  txtDec = ".")
occs_Mm <- userOccs_Mm$Mephitis_mephitis$cleaned
```

### Obtain environmental data

Using user-specified variables.

```{r}
## Specify the directory with the environmental variables
dir_envs_Mm <- ""
envs_path <- file.path(dir_envs_Mm, c('cal_veg_270m_wgs84_NEW_Clip.tif', 'dem_270m_wgs84_Clip.tif', 'distance_water_270m_wgs84_Clip.tif', 'mean_ppt_270m_wgs84_Clip.tif', 'mean_temp_270m_wgs84_Clip.tif', 'season_ppt_270m_wgs84_Clip.tif', 'season_temp_270m_wgs84_Clip.tif'))
# Create environmental object 
envs_Mm <- envs_userEnvs(
  rasPath = envs_path,
  rasName = c('cal_veg_270m_wgs84_NEW_Clip.tif', 'dem_270m_wgs84_Clip.tif', 'distance_water_270m_wgs84_Clip.tif', 'mean_ppt_270m_wgs84_Clip.tif', 'mean_temp_270m_wgs84_Clip.tif', 'season_ppt_270m_wgs84_Clip.tif', 'season_temp_270m_wgs84_Clip.tif'),
  doBrick = FALSE)
occs_xy_Mm <- occs_Mm[c('longitude', 'latitude')]
occs_vals_Mm <- as.data.frame(raster::extract(envs_Mm, occs_xy_Mm, cellnumbers = TRUE))
# Remove duplicated same cell values
occs_Mm <- occs_Mm[!duplicated(occs_vals_Mm[, 1]), ]
occs_vals_Mm <- occs_vals_Mm[!duplicated(occs_vals_Mm[, 1]), -1]
# remove occurrence records with NA environmental values
occs_Mm <- occs_Mm[!(rowSums(is.na(occs_vals_Mm)) >= 1), ]
# also remove variable value rows with NA environmental values
occs_vals_Mm <- na.omit(occs_vals_Mm)
# add columns for env variable values for each occurrence record
occs_Mm <- cbind(occs_Mm, occs_vals_Mm)
```

### Process Occurrence Data

Thinning the occurrences to 1 km

```{r}
# Thin occurrences 
occs_Mm <- poccs_thinOccs(
  occs = occs_Mm, 
  thinDist = 1)
```

### Process environmental data

Sampling of 10000 background points and corresponding environmental data
using a “minimum convex polygon” method with a 0.5 degree buffer.

```{r}
# Generate background extent 
bgExt_Mm <- penvs_bgExtent(
  occs = occs_Mm,
  bgSel = "minimum convex polygon",
  bgBuf = 0.5)
# Mask environmental data to provided extent
bgMask_Mm <- penvs_bgMask(
  occs = occs_Mm,
  envs = envs_Mm,
  bgExt = bgExt_Mm)
# Sample background points from the provided area
bgSample_Mm <- penvs_bgSample(
  occs = occs_Mm,
  bgMask =  bgMask_Mm,
  bgPtsNum = 10000)
# Extract values of environmental layers for each background point
bgEnvsVals_Mm <- as.data.frame(raster::extract(bgMask_Mm,  bgSample_Mm))
##Add extracted values to background points table
bgEnvsVals_Mm <- cbind(scientific_name = paste0("bg_", "Mephitis mephitis"), bgSample_Mm,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Mm)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using “hierarchical checkerboard”, a spatial partition method
with an aggregation factor of 2.

```{r}
# R code to get partitioned data
groups_Mm <- part_partitionOccs(
  occs = occs_Mm ,
  bg =  bgSample_Mm, 
  method = "cb2",
  bgMask = bgMask_Mm,
  aggFact = 2) 
```

### Build and Evaluate Niche Model

Generating a species distribution model using the maxnet algorithm as
implemented in ENMeval V2.0 (with clamping = TRUE). For tuning using LQ
feature classes and regularization multipliers in the 0.5, 0.5 range
increasing by 1. Not using any categorical predictor variables.

```{r}
# Run maxent model for the selected species
model_Mm <- model_maxent(
  occs = occs_Mm,
  bg = bgEnvsVals_Mm,
  user.grp = groups_Mm, 
  bgMsk = bgMask_Mm,
  rms = c(0.5, 0.5), 
  rmsStep =  1,
  fcs = 'LQ',
  clampSel = TRUE,
  algMaxent = "maxnet",
  parallel = FALSE,
  numCores = 11)
```

### Visualize

Generate a map of the maxnet generated model with no threshold

```{r}
# Select current model and obtain raster prediction
m_Mm <- model_Mm@models[["fc.LQ_rm.0.5"]]
predSel_Mm <- predictMaxnet(m_Mm, bgMask_Mm,
                                          type = "cloglog", 
                                          clamp = TRUE)
#Get values of prediction
mapPredVals_Mm <- getRasterVals(predSel_Mm, "cloglog")
#Define colors and legend  
rasCols <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
legendPal <- colorNumeric(rev(rasCols), mapPredVals_Mm, na.color = 'transparent')
rasPal <- colorNumeric(rasCols, mapPredVals_Mm, na.color = 'transparent')
#Generate map
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) 
m  %>%
  leaflet::addLegend("bottomright", pal = legendPal,
            title = "Predicted Suitability<br>(Training)",
            values = mapPredVals_Mm, layerId = "train",
            labFormat = reverseLabel(2, reverse_order = TRUE)) %>% 
  #add occurrence data
  addCircleMarkers(data = occs_Mm, lat = ~latitude, lng = ~longitude,
                   radius = 5, color = 'red', fill = TRUE, fillColor = "red",
                   fillOpacity = 0.2, weight = 2, popup = ~pop) %>% 
  ##Add model prediction
  addRasterImage(predSel_Mm, colors = rasPal, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
 ##add background polygons
  addPolygons(data = bgExt_Mm,fill = FALSE,
              weight = 4, color = "blue", group = 'proj')
```

### Visualize

Visualize response curves from “maxnet” model.

```{r}
# Retrieve env variables
n <- mxNonzeroCoefs(model_Mm@models[["fc.LQ_rm.0.5"]], "maxnet")

# Create response curves
for (i in n) {
maxnet::response.plot(
  model_Mm@models[["fc.LQ_rm.0.5"]],
  v = i,
  type = "cloglog")
}
```
